PACKER: An Exemplar Model of Category Generation

### 2017-01-29 Manuscripts/cogsci-2017/paper.tex
nbc pass back to joe

\begin{abstract}
FILLER (make very short)

\textbf{Keywords:} 
Categorization; etc
\end{abstract}

@@

One of the most intriguing capabilities of human cognition is the ability to generate new ideas and concepts. The creative creation of knowledge is an infrequent subject of formal inquiry. This paper focuses on one topic within creative cognition, category generation. Foundational work \citep[e.g.,][]{smith1993constraining,ward2002role,ward1994structured,ward1995s} found that people tend to create new categories that are similar to related previously known categories: New exemplars from the new categories tend to be copied from exemplars for a similar categories, but tweaked on one or a few dimensions. Thus, people generate a new category containing similar distributional properties to the categories it is based on, while being different from those categories in some manner.

Existing work on the creative use of conceptual knowledge  explores the role of prior knowledge in generating novel concepts. A core phenomenon is that people generate new categories with similar distributional properties as existing categories. For example, \citet{ward1994structured} asked participants to draw and describe plants and animals that might exist on other planets. Generation was strongly constrained by prior knowledge of Earth plants and animals: People generated alien species with the same features as those found on Earth (e.g., eyes, legs, wings) and possessing the same feature correlations observed on Earth (e.g., feathers co-occur with wings).

Recent work has proposed and tested computational models of how new categories are generated based on prior knowledge \citep{jern2013probabilistic}. In their experiments, they trained participants on a set of experimenter-defined categories composed of exemplars within an artificial three-dimensional domain (e.g., 2D shapes varying in size, hue, and saturation). After a short training phase, participants were asked to generate exemplars from a new category. Participants were provided with a set of scales to adjust the feature values of each generated stimulus, and were given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that people generated artificial categories possessing the same feature variance and correlations as the experimenter-defined categories in the domain. They evaluated different possible computational models of how people generate categories (such as "copy-and-tweak" and exemplar accounts) and found that a hierarchical Bayesian model generating new categories by abstracting the distributional structure of the categories in the domain.

%JLA: Add (a), (b) and (c) to figure and then use those to refer to each subpanel in the caption. It'll save space and be cleaner.
\begin{figure*}
    \begin{center}
    \inputpgf{figs/}{example-prob-spaces.pgf}
    \caption{PACKER generation of a category `B' example, following exposure to one member of category `A' and one member of category `B'. \textit{Left}: Predictions given $\{\phi = -1$, $\gamma = 0\}$ (contrast influence only). \textit{Center}: Predictions given $\{\phi = 0$, $\gamma = 1\}$ (target influence only).  \textit{Right}: Predictions given $\{\phi = -1$, $\gamma = 1\}$ (both constraints considered).  }
    \label{fig:example-prob-spaces}
    \end{center}
\end{figure*}

In this paper, we introduce a novel exemplar-based approach to category generation, the PACKER model (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}), which creates new categories by balancing two constraints: (1) new categories should be different from known categories (minimizing between-class similarity), and (2) new categories should be internally coherent (maximizing within-class similarity). As such, PACKER is a significant departure from previous accounts of generation -- rather than proposing that people create new categories that are mostly similar to related categories, PACKER creates new categories that are different from related categories. Further, it does so using the well-studied mechanics of exemplar representations and therefore possesses a rich connection to the wider body of research on category learning.

@@

The PACKER model is an extension of the prominent Generalized Context Model of category learning (GCM; \citealp{nosofsky1984abbr}). The model assumes that each category is represented by a collection of exemplars within in a $k$-dimensional psychological space, and that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories.

@@

To generate a new example given examples from contrast and target categories, the model considers both the similarity to examples from the contrast category as well as the similarity to examples in the target category. The aggregated similarity $a$ between generation candidate $y$ and the model's stored exemplars $x$ can be computed as $a(y, x) = \sum_j{f(x_j) s(y, x_j)}$, where $f(x_j)$ is a function specifying the extent to which each exemplar contributes to the target category. For brevity, we will abbreviate $f(x_j)$ as $f_j$. PACKER sets $f_j$ depending on exemplar $j$'s category membership: $f_j = \phi$ if $x_j$ is a member of the contrast category, and $f_j = \gamma$ if $x_j$ is a member of the target category. $\phi$ and $\gamma$ are free parameters ($-\infty \leq \phi, \gamma \leq \infty$) controlling the contribution of contrast- and target-category similarity, respectively. Larger absolute values for either parameter produce greater consideration of that exemplar for generating the target category, with values of 0 producing no effect. A negative value for $f_j$ produces a `repelling' effect (target exemplars are less likely to be generated nearby $x_j$). Conversely, a positive value for $f_j$ produces a `pulling' effect (target exemplars are more likely to be generated nearby $x_j$).

Recall that we proposed that categories are generated by balancing two constraints: that new categories should be different from existing categories, and exemplars belonging to the same category should be similar to one another. This is encoded in PACKER when $\phi < 0$, and $\gamma > 0$. Negative values for $\phi$ encourages $y$ to be distant from the contrast category (as similarity to contrast category exemplars are subtracted during aggregation). Positive values for $\gamma$ encourage $y$ to be close to other exemplars of the target category. When $|\phi| = \gamma$, the "force" of repulsion from the contrast category is equal to the "force" of attraction to the target category.

@@

Previous models of category generation focus on capturing the tendency for people to produce new categories that have similar distributional properties to existing categories. To the best of our knowledge, \citet{jern2013probabilistic} were the first to analyze behavioral results quantitatively to computational models. Based on their work, we describe two alternative models: a formalization of the \textit{copy-and-tweak} hypothesis \citep{ward2002role,ward1995s}, and the \textit{hierarchical sampling} proposed by \citet{jern2013probabilistic}.

\subsection{Copy-and-Tweak}
%JLA: PLEASE CHECK WHAT I WROTE TO MAKE SURE IT"S TRUE OF THEIR MODEL 
The \textit{copy-and-tweak} model proposes that generation is a two-part process: First, learners retrieve an observation from memory, and then they tweak it to be different from the retrieved observation. \citet{jern2013probabilistic} interpreted this proposal in terms of an exemplar model, where participants score potential new exemplars as the summed exponentiated similarity of the new exemplar to every exemplar from the category, and sample from the normalized score distribution. Given that this was intended for generating new exemplars within known category, we implemented a different formalization of the copy-and-tweak hypothesis.

In our implementation of the copy-and-tweak model, the probability that a given exemplar $z$ is retrieved from memory is $p(z) = \exp(f_z) / \sum_z{ \exp(f_z) }$, where $f_z$ is a function specifying each item's relative chance of being selected. In our simulations, we set $f_z$ to be $-\lambda$ for members of contrast categories, and $f_z = \lambda$ for members of the target category. The resulting free parameter $\lambda$ ($>0$) thus controls the relative chances that a member of the target category will be retrieved as the source. 
%\citet{jern2013probabilistic} did not implement any mechanisms to differentially weigh retrieval probabilities, and so our model encompasses theirs as a special case when $\lambda = 0$ (uniform probabilities). JLA: (i reread their model and i don't think this is quite true)

After a source exemplar is retrieved, the similarity between generation candidates $y$ and the retrieved exemplar is computed as per Equation \ref{eq:similarity}. The model's goal is to generate an item that is similar to $z$, but does not exceed a given similarity threshold: The generated item should be similar, but not too similar to $z$. Formally, the probability that candidate exemplar $y$ will be generated based on a source exemplar $z$ is

\begin{equation}
    p(y|z)  = \dfrac
    { \exp \left\{\theta \cdot s(y,z) \right\} I\left(s(y,z) \leq \tau\right) }
    {\sum_i{\exp \left\{ \theta \cdot s(y_i,z) \right\} I\left(s(y_i,z) \leq \tau\right)}} 
\end{equation}
% 
where $\theta$ is a response determinism parameter, $\tau$ is the similarity threshold, and $I(\cdot)$ is the indicator function, which returns 1 when it is passed a true expression and 0 otherwise. So, $I\left(s(y,z) \leq \tau\right)$ is 1 when the candidate exemplar $y$ is far enough away from the source exemplar $z$. This ensures the source exemplar is tweaked "enough". When $\tau=1$, the threshold has no effect, which results in a model similar to \citet{jern2013probabilistic}. To obtain predictions not depending on a given source example, the model's predictions can be aggregated over all possible sources: $p(y) = \sum_z{p(z)p(y|z)}$.
%Thus, the probability that $y$ will be selected is a function of its generation probability given each of the sources, and the probability each source will be retrieved.

@@

We assume that each category's exemplars are distributed according to a multivariate normal distribution with parameters ($\mu, \Sigma$).\footnote{Following \citet{jern2013probabilistic} stimuli were put through a logistic function to constrain the domain to be between 0 and 1.} We assume that each category's covariance matrix is inverse-Wishart distributed with parameters ($v$, $\kappa,$ and $\Sigma_D$).\footnote{Note that \citet{jern2013probabilistic}'s model is slightly different, as they used a non-conjugate model. Their model acts very similar to our version of it and receives comparable fits.} $\Sigma_D$ is the covariance matrix shared between categories. We assumed the shared covariance matrix $\Sigma_D$ was generated from a Wishart distribution (for conjugacy) with parameters $v_0$, $\kappa_0$, and $\Sigma_0$. We set $\nu_0 = \kappa_0 = 1$, and $\Sigma_0 = \rho {\bf I}$, where $\rho$ was a free parameter controlling the expected variance of dimensions (dimensions of the shared covariance matrix are expected to be uncorrelated) and ${\bf I}$ is the identity matrix.

To simplify the model predictions, we used {\em maximum a posteriori} (MAP) estimates for the hidden parameters and then generated new categories based on those estimates. Due to conjugacy, the MAP estimate for the shared covariance matrix $\Sigma_D = \Sigma_0 + \sum_c{C_c}$, where $C_c$ is the empirical covariance matrix of category $c$. The MAP estimate of the covariance matrix for category $B$ is 
\begin{equation}
  \Sigma_B = \left[ \Sigma_D \nu + C_B +
    \dfrac
    {\kappa n_B}
    {\kappa + n_B}
    (\bar{x_B}-\mu_B)(\bar{x_B}-\mu_B)^T
  \right] (\nu + n_B)^{-1}
  \label{eq:Sigma_B}
\end{equation}
%
where $\nu$ ($\nu>k-1$) is an additional free parameter (from the Inverse-Wishart prior on $\Sigma_B$) weighting the importance of $\Sigma_{D}$. When the target category has no members, $\Sigma_B = \Sigma_D$.

Generated exemplars are then assumed to be drawn from the multivariate normal distribution specified by $(\mu_{B}, \Sigma_{B})$. Thus, $p(y)$ is
\begin{equation}
  p(y) = \dfrac
    {\exp( \theta \cdot {\rm Normal}(y; \mu_{B}, \Sigma_{B}))}
    {\sum_i \exp( \theta \cdot {\rm Normal}(y_i; \mu_{B}, \Sigma_{B}))} 
\end{equation}
where $\theta$ is a response determinism parameter and ${\rm Normal}(y; \mu, \Sigma)$ denotes a multivariate Normal density evaluated at $y$.
%JLA: There's something funky going on with the notation. I got confused by your noation, so hopefully you can update what i did to be consistent with the rest.
\section{Behavioral Experiment}
A novel prediction of PACKER is that the contrast category's location within stimulus space should influence the generated category -- the generated category should be repulsed from the contrast category. The other models predict the location of the contrast category within stimulus space should not influence the generated category. This experiment tests this prediction and quantitatively evaluates the three models.

@@

The behavioral experiment described below was designed to test this key prediction:  Does the \textit{location} of contrast categories, influence generation? The experiment follows the paradigm developed by \citet{jern2013probabilistic}: first, participants learn members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B'). We developed two Alpha categories (see Figure \ref{fig:middle-bottom-conditions}). We did so by creating one of the Bottom Alpha category as a tight cluster and formed the Middle Alpha category by translating it to the center of stimulus space. As the two categories only differ in their location, the two categories should not differ in their distributional information.

@@

@@

\begin{figure}[ht!]
    \begin{center}
    \inputpgf{figs/}{beta.samples.pgf}
    \caption{Sample generated categories. }
    \label{fig:beta.samples}
    \end{center}
\end{figure}

@@

Because the conditions differ only in their location along the y-axis (see Figure \ref{fig:middle-bottom-conditions}), we focus on how Beta exemplars are generated above and below the contrast category. Several sample Beta categories are depicted in Figure \ref{fig:beta.samples}. As is evident in Figure \ref{fig:beta.samples}, we observed large individual differences in generation strategy: Whereas some participants generated all four Beta examples within a narrow y-axis range, others generated Beta examples along a wide range.

@@

\begin{table}
\begin{center} 
\label{table:subset-table} 
\vskip 0.12in
\begin{tabular}{ l r r}
    \textbf{Middle}         & Used top row & No top row \\
    \hline
    Used bottom rows       &  28 & 18  \\
    No bottom row          &  11 &  4  \\
    \\
    \textbf{Bottom}         & Used top row & No top row \\
    \hline
    Used bottom row        & 16 & 8 \\
    No bottom row          & 31 & 6 \\
\end{tabular}
\caption{Behavioral results.} 
\end{center} 
\end{table}
To evaluate PACKER's other predictions, we computed the number of exemplars produced at different distances to the center of the Alpha category. These data, depicted in the top panel of Figure \ref{fig:distance.figs} reveal a strong preference for stimuli that are dissimilar to the members of the Alpha category: maximally distant items were by far the most frequently generated.

Finally, we computed for each participant the average distance between exemplars belonging to the same and opposite categories. These data (depicted in the bottom panel of Figure \ref{fig:distance.figs}) show that, as observed by \citet{ward1994structured}, most people generated Beta categories in which members are closer to one another than they are to members of the Alpha category (i.e., more between- than within-category distance). We did however, observe a notable subset of individuals with greater within-class distance. These individuals tended to adopt a `corners' approach, in which a Beta examples were placed almost exclusively in the corners of the space.

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{distance.figs.pgf}
    \caption{Behavioral results. \textit{Top}: Frequency of exemplar generation as a function of distance from the Alpha category normalized by the maximum possible distance. \textit{Bottom}: Within- vs. between-category distance for every participant. }
    \label{fig:distance.figs}
    \end{center}
\end{figure}

\subsection{Summary}
Our results support PACKER's predictions: People tend to generate items that are dissimilar from the contrast category Alpha and similar to the target category Beta (with some exceptions, see Figure \ref{fig:distance.figs}). By consequence, we observed considerable differences in generation between the Middle and Bottom conditions: participants in the Bottom condition were less likely to use the bottom row of the stimulus space for generation, and participants in the Middle condition were more likely to create categories spanning the entire y-axis (utilizing the top and bottom row of the space). This latter result is especially interesting as it conflicts with previous results: Qualitatively different types of categories are generated (with very different distributional information), depending on only the location of the Alphas in the domain.

@@

\section{Model Evaluation}
In this section, we evaluate the model fits to our results. To obtain an overall sense of each model's ability to explain our results, we fit each model by maximizing the log-likelihood of the model's predictions of the human results. Four free parameters were fitted in each model. The $c$, $\phi$, $\gamma$, and $\theta$ parameters were fitted for PACKER; $c$, $\lambda$, $\tau$, and $\theta$ were fitted for the copy-and-tweak model, and $\kappa$, $\rho$, $\nu$, and $\theta$ were fitted for the hierarchical sampling model. Note that each model possess a $\theta$ parameter fulfilling the same role (response determinism), and PACKER shares with copy-and-tweak a specificity parameter, $c$. Attention in PACKER and copy-and-tweak was set uniformly. Parameters were not allowed to vary between participants or conditions -- the goal was to obtain the best-fitting values to our entire dataset.

Each model's best-fitting parameterization is shown in Table \ref{table:model-fits}. Overall, PACKER outperformed copy-and-tweak and the hierarchical sampling model by a considerable margin ($\sim17\%$ improvement in log-likelihood). The parameter settings associated with PACKER's best fit are exactly as expected: a strong preference for items that are similar to members of the target category but are dissimilar to members of the contrast category. A similar pattern of results was obtained when we only considered the second to fourth exemplars generated by each participant.
%JLA this is fine
\begin{table}[ht!]
\centering
\caption{Model-fitting results.}
\label{table:model-fits}
\begin{tabular}{ l l l}
\\
 \textbf{PACKER} & \textbf{Copy \& Tweak} & \textbf{Hierarchical} \\
 & & \textbf{Sampling} \\ \hline
 $L = -1733$ & $L = -2040$ & $L = -2055$  \\ 
 $c=0.564$ & $c=9.445$  & $\kappa<0.001$\\
$\phi=-13.608$& $\lambda=15.115$ & $\rho=1.001$ \\ 
$\gamma=12.006$  & $\tau=0.415$ & $\nu=0.451$  \\ 
 $\theta=0.259$  & $\theta=7.068$ & $\theta=6.137$  \\ 
\end{tabular}
\end{table}

The model-fitting results above follow directly from the design of our experiment. Because distributional structure is the primary determinant of generation in the copy-and-tweak and hierarchical sampling models, those models have no basis to predict any differences between Middle and Bottom conditions (which do not differ in distributional structure).

\subsection{Individual Differences}
%In future work, we intend to explore more systematically the role of PACKER's parameters in different subtypes of performance. For example, whereas the tight cluster profile could be explained with a moderately negative $\phi$ and a strongly positive $\gamma$, the four corners profile could be simulated with a strongly negative $\phi$ and $\gamma\leq0$.
%While it is important that any model be able to explain different types of performance with the same parameter values, these radical differences in generation are difficult to account for under a single set of parameters.
Our behavioral results showed that, in general, generated categories consist of item that are dissimilar to members of opposite categories, and similar to members of the their own category. There were four typical patterns of Beta categories: tightly clustered, four corners, `column'-like categories (varying along the y-axis but not the x-axis), others created `row'-like categories (see Figures \ref{fig:beta.samples} and \ref{fig:distance.figs}). Here, we explore the role of PACKER's different attention weights ($w_1$ and $w_2$) in producing row-like or column-like categories.

%JLA: Move PACKER to be between Behavioral and Copy and Tweak. Also, I think this will take up less space as a row figure.
\begin{figure}[ht!]
    \begin{center}
    \inputpgf{figs/}{range-diff-gradient.pgf}
    \caption{Generated category structure as a function of location. Orange areas in each gradient correspond to stimuli that were commonly generated into category possessing greater y-axis range (columns). Purple areas correspond to categories possessing greater x-axis range. White areas correspond to equal range along both features (or infrequent generation).}
    \label{fig:range-diff-gradient}
    \end{center}
\end{figure}

>>>>

\begin{abstract}

The creative generation of new concepts is an intriguing yet understudied topic in cognitive science. In this paper, we present a novel exemplar-based approach to category generation: PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}). PACKER's core design assumptions are that (\textit{a}) categories are represented as exemplars in a multidimensional psychological space, (\textit{b}) generated items should be similar to exemplars of the same category, and (\textit{c}) generated categories should be dissimilar to existing categories. A behavioral study and formal simulations reveal broad support for PACKER's predictions.

\textbf{Keywords:} 
Categorization, exemplar models, category generation, creative cognition.
\end{abstract}

@@

The creation of new concepts and ideas is among the most interesting -- yet infrequently studied -- capabilities of human cognition. This paper focuses on one topic within the broader field of creative cognition: category generation. Foundational work on this topic \citep[e.g.,][]{smith1993constraining,ward2002role,ward1994structured,ward1995s} has focused on the role of prior knowledge in generating novel concepts. A core phenomenon is that people generate categories with similar distributional properties as existing categories. For example, \citet{ward1994structured} asked participants to generate species of plants and animals that might exist on other planets. Generation was strongly constrained by prior knowledge of Earth species: People generated species with the same features as those found on Earth (e.g., eyes, legs, wings) and possessing the same feature correlations observed on Earth (e.g., feathers co-occur with wings).

Recent work has proposed and tested computational models of the processes underlying category generation. \citet{jern2013probabilistic} trained participants on a set of experimenter-defined categories composed of exemplars within an artificial three-dimensional domain (e.g., 2D shapes varying in size, hue, and saturation). After a short training phase, participants were asked to generate exemplars from a new category. Participants were provided with a set of scales to adjust the feature values of each generated stimulus, and were given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that generated categories possessed the same feature variance and correlations as the experimenter-defined categories in the domain. They evaluated different possible computational models of how people generate categories (such as an exemplar-based "copy-and-tweak" model) and found that a hierarchical Bayesian model (which generates categories by abstracting the distributional structure of the domain) provided the strongest account of the behavioral results.

\begin{figure*}
    \begin{center}
    \inputpgf{figs/}{example.spaces.pgf}
    \caption{PACKER generation of a category `B' example, following exposure to one member of category `A' and one member of category `B'. \textit{(a)} Predictions based on contrast similarity only $\{\phi = -1$, $\gamma = 0\}$. \textit{(b)} Predictions based on target similarity only $\{\phi = 0$, $\gamma = 1\}$.  \textit{(c)} Predictions with both constraints considered $\{\phi = -1$, $\gamma = 1\}$.  }
    \label{fig:example-prob-spaces}
    \end{center}
\end{figure*}

In this paper, we introduce a novel exemplar-based approach to category generation, PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}), which creates  categories by balancing two constraints: (1) new categories should be different from known categories (minimizing between-class similarity), and (2) new categories should be internally coherent (maximizing within-class similarity). As such, PACKER is a significant departure from previous accounts of generation -- rather than proposing that people create categories by abstracting and re-using knowledge of related categories, PACKER first considers how the generated category should \textit{differ} from related categories. Further, it does so using the well-studied mechanics of exemplar representations and therefore possesses a rich connection to the wider body of research on category learning.

@@

The PACKER model is an extension of the prominent Generalized Context Model of category learning (GCM; \citealp{nosofsky1984choice}). The model assumes that each category is represented by a collection of exemplars within in a $k$-dimensional psychological space, and that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories.

@@

To generate a new example, the model considers both the similarity to examples from contrast categories as well as similarity to examples in the target category. The aggregated similarity $a$ between generation candidate $y$ and the model's stored exemplars $x$ can be computed as $a(y, x) = \sum_j{f(x_j) s(y, x_j)}$, where $f(x_j)$ is a function specifying the extent to which each exemplar contributes to the target category. For brevity, we will abbreviate $f(x_j)$ as $f_j$. PACKER sets $f_j$ depending on exemplar $j$'s category membership: $f_j = \phi$ if $x_j$ is a member of the contrast category, and $f_j = \gamma$ if $x_j$ is a member of the target category. $\phi$ and $\gamma$ are free parameters ($-\infty \leq \phi, \gamma \leq \infty$) controlling the contribution of contrast- and target-category similarity, respectively. Larger absolute values for either parameter produce greater consideration of those exemplars, with values of 0 producing no effect. A negative value for $f_j$ produces a `repelling' effect (exemplars are less likely to be generated nearby $x_j$). Conversely, a positive value for $f_j$ produces a `pulling' effect (exemplars are more likely to be generated nearby $x_j$).

Recall that we proposed that categories are generated by balancing two constraints: that new categories should be different from existing categories, and exemplars belonging to the same category should be similar to one another. This is realized in PACKER when $\phi < 0$, and $\gamma > 0$. Negative values for $\phi$ encourage $y$ to be distant from the contrast category (as similarity to contrast category exemplars are subtracted during aggregation). Positive values for $\gamma$ encourage $y$ to be close to other exemplars of the target category. When $|\phi| = \gamma$, the "force" of repulsion from the contrast category is equal to the "force" of attraction to the target category.

@@

Previous models of category generation focus on capturing the tendency for people to produce new categories that have similar distributional properties to existing categories. To the best of our knowledge, \citet{jern2013probabilistic} were the first to quantitatively evaluate computational models of generation. Based on their work, we describe two alternative models: a formalization of the \textit{copy-and-tweak} hypothesis \citep{ward2002role,ward1995s}, and the \textit{hierarchical sampling} proposed by \citet{jern2013probabilistic}.

\subsection{Copy-and-Tweak}
The \textit{copy-and-tweak} model proposes that generation is a two-part process: First, learners retrieve an observation of the target class from memory, and then they tweak it make something new. \citet{jern2013probabilistic} interpreted this proposal in terms of an exemplar model using the GCM \citep{nosofsky1984choice}. Formally, their model is equivalent to PACKER with $\{\phi = 0$, $\gamma = 1\}$ (see Figure \ref{fig:example-prob-spaces}). To allow for more flexible performance, and to explicitly formalize the `tweak' procedure, we slightly generalized their implementation.

In our copy-and-tweak model, all observations $z$ from the target class are equally likely to be retrieved as the source: $p(z) = 1 / n$, where $n$ is the number of stored examples in the target class. If $n=0$, learners are assumed to generate an item at random. After a source exemplar is retrieved, the similarity between generation candidates $y$ and the retrieved exemplar is computed as per Equation \ref{eq:similarity}. The model's goal is to generate an item that is similar to $z$, but does not exceed a given similarity threshold: The generated item should be similar, but not too similar to $z$. Formally, the probability that candidate exemplar $y$ will be generated based on a source exemplar $z$ is

\begin{equation}
    p(y|z)  = \dfrac
    { \exp \left\{\theta \cdot s(y,z) \right\} I\left(s(y,z) \leq \tau\right) }
    {\sum_i{\exp \left\{ \theta \cdot s(y_i,z) \right\} I\left(s(y_i,z) \leq \tau\right)}} 
\end{equation}
% 
where $\theta$ is a response determinism parameter, $\tau$ is the similarity threshold, and $I(\cdot)$ is the indicator function, which returns 1 when it is passed a true expression and 0 otherwise. So, $I\left(s(y,z) \leq \tau\right)$ is 1 when the candidate exemplar $y$ is far enough away from the source exemplar $z$. This ensures the source exemplar is tweaked "enough". When $\tau=1$, the threshold has no effect, and our model is equivalent to that of \citet{jern2013probabilistic}. To obtain predictions not depending on a given source example, the model's predictions can be aggregated over all possible sources: $p(y) = \sum_z{p(z)p(y|z)}$.

@@

We assume that each category's exemplars are distributed according to a multivariate normal distribution with parameters ($\mu, \Sigma$). We assume that each category's covariance matrix is inverse-Wishart distributed with parameters ($v$, $\kappa,$ and $\Sigma_D$).\footnote{Note that \citet{jern2013probabilistic}'s model is slightly different, as they used a non-conjugate model. Their model acts very similar to our version of it and receives comparable fits.} $\Sigma_D$ is the covariance matrix shared between categories. We assumed the shared covariance matrix $\Sigma_D$ was generated from a Wishart distribution (for conjugacy) with parameters $v_0$, $\kappa_0$, and $\Sigma_0$. We set $\nu_0 = \kappa_0 = 1$, and $\Sigma_0 = \rho {\bf I}$, where $\rho$ was a free parameter controlling the expected variance of dimensions (dimensions of the shared covariance matrix are expected to be uncorrelated) and ${\bf I}$ is the identity matrix.

To simplify the model predictions, we used {\em maximum a posteriori} (MAP) estimates for the hidden parameters and then generated new categories based on those estimates. Due to conjugacy, the MAP estimate for the shared covariance matrix $\Sigma_D = \Sigma_0 + \sum_c{C_c}$, where $C_c$ is the empirical covariance matrix of category $c$. The MAP estimate of the covariance matrix for category $B$ is 
\begin{equation}
  \Sigma_B = \left[ \Sigma_D \nu + C_B +
    \dfrac
    {\kappa n_B}
    {\kappa + n_B}
    (\bar{x_B}-\mu_B)(\bar{x_B}-\mu_B)^T
  \right] (\nu + n_B)^{-1}
  \label{eq:Sigma_B}
\end{equation}
%
where $\nu$ ($\nu>k-1$) is an additional free parameter (from the Inverse-Wishart prior on $\Sigma_B$) weighting the importance of $\Sigma_{D}$. When the target category has no members (i.e., $n_B = 0$), items are generated at random.

Generated exemplars are then assumed to be drawn from the multivariate normal distribution specified by $(\mu_{B}, \Sigma_{B})$. Thus, $p(y)$ is
\begin{equation}
  p(y) = \dfrac
    {\exp( \theta \cdot {\rm Normal}(y; \mu_{B}, \Sigma_{B}))}
    {\sum_i \exp( \theta \cdot {\rm Normal}(y_i; \mu_{B}, \Sigma_{B}))} 
\end{equation}
where $\theta$ is a response determinism parameter and ${\rm Normal}(y; \mu, \Sigma)$ denotes a multivariate Normal density evaluated at $y$. 
%JLA: There's something funky going on with the notation. I got confused by your noation, so hopefully you can update what i did to be consistent with the rest.

\section{Behavioral Experiment}

The copy-and-tweak and hierarchical sampling models were designed to explain effects of prior knowledge on the structure of categories, and so these models assume little about where in stimulus space categories are likely to be generated. Indeed, when there are no known examples of the target category, both models assume that generation is random. PACKER is thus unique in its prediction that contrast categories should influence both the structure and location of generated categories. The behavioral experiment described below was designed to test this key prediction.

The experiment follows the paradigm developed by \citet{jern2013probabilistic}: first, participants learn members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B'). We developed two Alpha categories (see Figure \ref{fig:middle-bottom-conditions}): the `Bottom' Alpha category is a tight cluster in the bottom-center of the space, and the `Middle' Alpha category is identical except that it lies in the center of stimulus space. The two categories only differ in their location, not in  distributional structure.

@@

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{beta.samples.pgf}
    \caption{Sample generated categories. }
    \label{fig:beta.samples}
    \end{center}
\end{figure}

@@

@@

Because the conditions differ only in their location along the y-axis (see Figure \ref{fig:middle-bottom-conditions}), we focus on how Beta exemplars are generated above and below the contrast category. Several sample Beta categories are depicted in Figure \ref{fig:beta.samples}. As is evident in Figure \ref{fig:beta.samples}, we observed broad individual differences in generation strategy: Whereas some participants generated all four Beta examples within a narrow y-axis range, others generated Beta examples along a wide range.

@@

\begin{table}
\begin{center} 
\caption{Behavioral results.} 
\label{table:subset-table} 
\vskip 0.12in
\begin{tabular}{ l r r}
    \textbf{Middle}         & Used top row & No top row \\
    \hline
    Used bottom rows       &  28 & 18  \\
    No bottom row          &  11 &  4  \\
    \\
    \textbf{Bottom}         & Used top row & No top row \\
    \hline
    Used bottom row        & 16 & 8 \\
    No bottom row          & 31 & 6 \\
\end{tabular}
\end{center} 
\end{table}

To evaluate PACKER's other predictions, we computed the number of exemplars produced at different distances to the center of the Alpha category. These data, depicted in the left panel of Figure \ref{fig:distance.figs} reveal a strong preference for stimuli that are dissimilar to the members of the Alpha category: maximally distant items were by far the most frequently generated.

Finally, we computed for each participant the average distance between exemplars belonging to the same and opposite categories. These data (depicted in the right panel of Figure \ref{fig:distance.figs}) show that, as observed by \citet{ward1994structured}, most people generated Beta categories in which members are closer to one another than they are to members of the Alpha category (i.e., more between- than within-category distance). We did however, observe a notable subset of individuals with greater within-class distance. These individuals tended to adopt a `corners' approach, in which a Beta examples were placed almost exclusively in the corners of the space.

\subsection{Summary}

Our results support PACKER's predictions: People tend to generate items that are dissimilar from the contrast category and similar to the target category (with some exceptions, see Figure \ref{fig:distance.figs}). By consequence, we observed considerable differences in generation between the Middle and Bottom conditions: participants in the Bottom condition were less likely to use the bottom row of the stimulus space for generation, and participants in the Middle condition were more likely to create categories spanning the entire y-axis (utilizing the top and bottom row of the space). This latter result is especially interesting as it conflicts with previous results: Qualitatively different types of categories are generated (with very different distributional information), depending on only the location of the Alphas in the domain.

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{distance.figs.pgf}
    \caption{Behavioral results. \textit{Left}: Frequency of exemplar generation as a function of distance from the Alpha category normalized by the maximum possible distance. \textit{Right}: Within- vs. between-category distance for every participant. }
    \label{fig:distance.figs}
    \end{center}
\end{figure}

\section{Model Evaluation}
To obtain an overall sense of each model's ability to explain our results, we fit each model by maximizing the log-likelihood of the model's predictions of the human results. The $c$, $\phi$, $\gamma$, and $\theta$ parameters were fitted for PACKER; $c$, $\tau$, and $\theta$ were fitted for the copy-and-tweak model, and $\kappa$, $\rho$, $\nu$, and $\theta$ were fitted for the hierarchical sampling model. Note that each model possess a $\theta$ parameter fulfilling the same role (response determinism), and PACKER shares with copy-and-tweak a specificity parameter, $c$. Attention in PACKER and copy-and-tweak was set uniformly. Parameters were not allowed to vary between participants or conditions -- the goal was to obtain the best-fitting values to our entire dataset.

Each model's best-fitting parameterization is shown in Table \ref{table:model-fits}. Overall, PACKER outperformed copy-and-tweak and the hierarchical sampling model by a considerable margin ($\sim17\%$ improvement in log-likelihood). The parameter settings associated with PACKER's best fit are exactly as expected: a strong preference for items that are similar to members of the target category but are dissimilar to members of the contrast category. A similar pattern of results was obtained when we only considered the second to fourth exemplars generated by each participant.

Our model-fitting results follow directly from the design of our experiment. Because the copy-and-tweak and hierarchical sampling models are not influenced by the location of contrast categories within the space, they do not capture the broad tendency for generated categories to be dissimilar to existing classes.

\begin{table}
\centering
\caption{Model-fitting results.}
\label{table:model-fits}
\begin{tabular}{ l l l}
\\
 \textbf{PACKER} & \textbf{Copy \& Tweak} & \textbf{Hierarchical} \\
 & & \textbf{Sampling} \\ \hline
 $AIC = 3476$ & $AIC = 3937$ & $AIC = 4061$  \\ 
 $c=0.565$ & $c=4.675$  & $\kappa<0.001$\\
$\phi=-4.814$ & $\tau=0.895$ & $\nu=1.001$ \\ 
$\gamma=4.250$ & $\theta=5.601$ &  $\rho=1.220$  \\ 
 $\theta=0.731$ &  & $\theta = 7.147$  \\ 
\end{tabular}
\end{table}

\subsection{Individual Differences}
%In future work, we intend to explore more systematically the role of PACKER's parameters in different subtypes of performance. For example, whereas the tight cluster profile could be explained with a moderately negative $\phi$ and a strongly positive $\gamma$, the four corners profile could be simulated with a strongly negative $\phi$ and $\gamma\leq0$.
%While it is important that any model be able to explain different types of performance with the same parameter values, these radical differences in generation are difficult to account for under a single set of parameters.
Our behavioral results showed that, in general, generated categories consist of items that are dissimilar to members of opposite categories, and similar to members of the their own category. However, beyond this broad description of performance, we observed a great deal of individual differences in generation style. Manually inspecting the data reveals four typical patterns (see Figure \ref{fig:beta.samples}): `corners' categories with one Beta example in each corner of the space, tight clusters, `column'-like categories varying along the y-axis but not the x-axis, and `row'-like categories varying along the x-axis but not the y-axis. Here, we explore the role of PACKER's attention weights ($w_1$ and $w_2$) in producing row-like or column-like categories.
